### 2025-04-06 Thesis/Latex/Chapter_2.tex
update

\subsection{Historical development of NLP}
The figure below presents a timeline outlining the key developments in Natural Language Processing (NLP)  beginning with rule-based approaches in the 1950s and progressing through the rise of statistical methods and early neural networks in the late 1980s. It then highlights the impact of deep learning from the 2000s onward, leading to the development of pre-trained models like BERT and GPT. The timeline concludes with the emergence of large language models (LLMs) from 2019 to the present, marking a significant shift in NLP research and applications.
\begin{figure}[htbp]
	\centerline{\includegraphics[scale=.7]{Figures/the_historyOf_nlp.png}}
	\caption{The Vauquois triangle, illustrating the foundations of machine translation.}
	\label{the_historyOf_nlp.png}
\end{figure}

@@

\subsection{Basics Concepts in Neural Networks}
\textbf{Weights} \\
Variables on the edges between nodes that are multiplied with node outputs to form the input for the next layer.Weights are crucial for training and tuning a neural network and are often initialized within a range of -1 to 1.  \\ \\ 
\textbf{Biases} \\
Additional nodes in hidden and output layers that connect to every node within their respective layers but not to the previous layer. Biases add a constant value (typically 1 or -1) to the input of a layer, helping shift the activation function and aiding in effective learning. 
\textbf{Activation Functions} \\
These functions introduce non-linearity to the network, enabling it to learn and model complex data patterns. Common activation functions include the sigmoid, hyperbolic tangent (tanh), and rectified linear unit (ReLU). \cite{taylor2017neural}.

\subsection{ Recurrent Neural Networks (RNNs)}
Recurrent neural networks (RNNs) are a type of neural network designed for processing sequential data, such as text or time series, by effectively handling variable-length inputs. An RNN consists of a hidden state h and an optical output y, which operate on a variable  sequence x=($x_1$....$x_t$)At each time step t, the hidden state $h^t$
of the RNN is updated according to the function $h^{(t)} = f(h^{(t-1)}, x_t)$
where f is a non-linear activation function, such as a logistic sigmoid or a more complex long short-term memory (LSTM) unit. This design allows RNNs to remember past inputs and incorporate them into future outputs, enabling them to recognize patterns that occur at multiple positions within a sequence. Additionally, RNNs utilize parameter sharing across time steps, which helps in generalizing across sequences of varying lengths and learning complex dependencies over time. Despite challenges with long-term dependencies, RNNs are particularly effective for tasks requiring context or sequential understanding, making them valuable for various applications, including time series analysis and natural language processing\cite{cho2014learning} 
\begin{figure}[htbp]

	\centerline{\includegraphics[width=1\linewidth]{Figures/RNN.png}}
	\caption{RNN Architecture.}
	\label{RNN.png}
\end{figure}
\subsection{ Long-Short Term Memory (LSTM)}
Long Short-Term Memory (LSTM) networks, designed by Hochreiter and Schmidhuber, are an improved version of recurrent neural networks (RNNs) designed to learn long-term dependencies in sequential data, making them suitable for tasks like time series forecasting and language translation
They address the limitations of traditional RNNs by introducing a memory cell that maintains information over extended periods. This memory cell is regulated by three gates—input, forget, and output gates—which control the flow of information in and out of the cell as shown in the fig \cite{geeksforgeeks_lstm}
\begin{figure}[htbp]
	\centering\includegraphics[width=0.9\linewidth]{Figures/LSTM.png}
	\caption{LSTM Architecture.}
	\label{/LSTM.png}
\end{figure}
Forget Gate: Decides what information to discard from the cell state
\[
f_t = \sigma \left( W_f \cdot [h_{t-1}, x_t] + b_f \right)
\]

Input Gate: Determines what new information to add to the cell state
\[
i_t = \sigma \left( W_i \cdot [h_{t-1}, x_t] + b_i \right)
\]
\[
\tilde{C}_t = \tanh \left( W_c \cdot [h_{t-1}, x_t] + b_c \right)
\]

Output Gate: Controls what information to output from the cell state
\[
o_t = \sigma \left( W_o \cdot [h_{t-1}, x_t] + b_o \right)
\]
\[
h_t = o_t \odot \tanh(C_t)
\]

It is worth mentioning that Bidirectional LSTM networks extend the traditional LSTM architecture by processing data in both forward and backward directions. This approach enables the network to capture dependencies from both past and future contexts, improving its ability to resolve temporal dependencies. Bidirectional LSTMs are particularly effective at handling multidimensional problems, encapsulating spatially and temporally distributed information, and dealing with incomplete data through flexible connection mechanisms \cite{hochreiter1997long}. 
\section{ Transformer Architecture}
The Transformer architecture is a deep learning model introduced in June 2017 by Vaswani et al. from Google Brain. Their paper, titled "Attention Is All You Need," presented a groundbreaking approach to processing sequential data through the use of a self-attention mechanism. This innovative method allows the model to assign different levels of importance to various parts of the input, enabling it to capture long-range dependencies much more effectively than earlier models like RNNs and LSTMs.The original Transformer model is structured as a stack of six layers, where the output of each layer i serves as the input to the subsequent layer i+1, continuing this process until the final prediction is reached.It features a six-layer encoder on the left and a corresponding six-layer decoder on the right,  both of which work together to transform input sequences into meaningful outputs. Each encoder and decoder consists of six identical layers that allow the model to process and generate language efficiently \cite{rothman2021transformers}.
\begin{figure}[htbp]

@@

\end{figure}
\subsection{Encoder and Decoder Stacks}
\textbf{Encoder:} Each layer in the encoder consists of two sub-layers: \\
Multi-head self-attention mechanism: This allows the model to focus on different parts of the input sequence. \\
Position-wise feed-forward network: A fully connected network applied to each position separately and identically. \\
Residual Connections: Residual connections are applied around each sub-layer, followed by layer normalization. The output of each sub-layer is computed as Layer Norm(x + Sublayer(x)). \\
Output Dimension: All sub-layers and embedding layers output vectors of dimension d-model = 512.\\
\textbf{Decoder:} Similar to the encoder, the decoder also has 6 identical layers, each containing the same two sub-layers:
Multi-head self-attention mechanism ,
Position-wise feed-forward network. \\
Additional Sub-layer: The decoder includes a third sub-layer that performs multi-head attention over the output of the encoder stack. \\
Residual Connections and Normalization: Like the encoder, residual connections are applied around each sub-layer, followed by layer normalization. \\
Masked Self-Attention: The self-attention mechanism in the decoder is modified to prevent positions from attending to subsequent positions, ensuring that predictions for position i depend only on the known outputs at positions before i \cite{vaswani2017attention}.
\begin{figure}[htbp]

@@

\end{figure}
\subsection{Self-Attention Mechanism}
Self-attention is a mechanism that allows the model to weigh the importance of different words in a sequence relative to each other. This is crucial for understanding context and relationships between words \cite{vaswani2017attention}.
\subsection{Query, Key, and Value Vectors:}
\textbf{Query Vectors (Q):} Generated from the input word's embedding and represent the word's role in finding relevant information in other words.\\
\textbf{Key Vectors (K):} Derived similarly from embedding and serve as a comparison standard for each word.\\
\textbf{Value Vectors (V):} Contain the actual information to be passed on and are weighted according to relevance determined by the Query and Key comparison \cite{vaswani2017attention}.. 
\subsection{Self-Attention Calculation:}
The score is computed by taking the dot product of the Query vector of a word with the Key vectors of all other words. This score determines how much focus should be placed on each word when encoding the current word.
The scores are scaled and passed through a softmax function, producing a distribution of attention weights. \\
Each Value vector is weighted by these attention scores and summed to produce the final output of the self-attention layer\cite{vaswani2017attention}.
\subsection{Role of Multi-Head Attention}
Multi-head attention enhances the model’s ability to focus on different aspects of the input by creating multiple sets of Query, Key, and Value vectors. Each set is processed independently and then combined. \\
This approach allows the model to attend to information from different representation subspaces at different positions, improving its ability to capture complex relationships\cite{vaswani2017attention}.
\subsection{Positional Encoding:}
Since self-attention mechanisms do not inherently capture the order of words, positional encoding is added to the word embeddings to provide information about the relative position of each word in the sequence.
These encodings are added directly to the input embeddings, allowing the model to consider both the position and the content of 
each word during processing\cite{alammar2018illustratedtransformer}.
\subsection{Transformers in NLP}
Transformers have revolutionized the field of Natural Language Processing (NLP) by introducing a powerful and efficient way to handle textual data. This advancement has led to the creation of highly effective models like BERT and GPT, with its deep bidirectional context, excels at understanding and improving performance across various NLP tasks. GPT, on the other hand, is renowned for generating coherent and contextually relevant text, significantly advancing applications such as text generation and translation\cite{devlin2019bert}.\\
\textbf{Scalability:} Transformers efficiently handle large datasets and long sequences, overcoming the limitations of RNNs. This scalability allows for training with billions of parameters, enhancing model capabilities.\\
\textbf{Rich Contextual Understanding: }The self-attention mechanism in transformers captures relationships between words across the entire sequence, enabling deep contextual understanding and more accurate language processing.\\
\textbf{Model Efficiency:} Transformers enable parallel processing, which speeds up training and makes them more efficient than sequential models like RNNs. This efficiency supports the rapid development and deployment of advanced language models\cite{vaswani2017attention}.
\section{ Emergence of Large Language Models (LLMs)}

Large Language Models (LLMs) are advanced artificial intelligence systems designed to process and generate human-like text. They are typically based on transformer architectures and are characterized by their enormous scale, with billions of parameters. LLMs are pre-trained on vast amounts of text data in a self-supervised manner, enabling them to develop a broad understanding of language. They are capable of performing a wide range of tasks with minimal task-specific fine-tuning, often achieving significant performance improvements through few-shot or zero-shot learning\cite{brown2020language}.
\subsection{Scaling in Large Language Models (LLMs)}
Scaling is crucial in the evolution of large language models. The history of scaling shows that increasing both model size and dataset size leads to significant improvements in performance across various NLP tasks. For instance, early work by Brants et al. (2007) demonstrated the benefits of using language models trained on vast datasets, such as 2 trillion tokens, which led to significant advancements in machine translation quality. This was followed by efforts like those of Heafield et al. (2013), who scaled traditional models to Web-scale data, and Jozefowicz et al. (2016), who scaled LSTMs to 1 billion parameters, achieving state-of-the-art results on large benchmarks.The advent of transformer-based models marked a significant shift.Models like BERT, GPT-2, and GPT-3, with their enormous parameter counts—up to 175 billion for GPT-3—demonstrated that scaling up not only the model but also the dataset size yields substantial gains in performance. Researchers like Kaplan et al. (2020) and Hoffmann et al. (2022) studied how scaling affects model performance, proposing power laws that show a predictable relationship between model size, dataset size, and performance. These studies emphasized the importance of scaling for the continued progress of LLMs\cite{touvron2023llama}.
\subsection{Pre-training}
Pre-training is a crucial stage in developing Large Language Models (LLMs), where the model learns from extensive unlabeled datasets through a process called self-supervision. This stage allows the model to recognize and internalize a wide range of linguistic patterns, laying the groundwork for fine-tuning on specific tasks.\\
Several pre-training objectives have been implemented to maximize the effectiveness of this learning process, each offering distinct benefits to the model's performance
\textbf{Full Language Modeling:} Used since GPT-2, this approach trains decoder-only models to predict the next token in a sequence based on previous tokens. This autoregressive method enables models like GPT-3 to generate coherent and contextually relevant text.\\
\textbf{Prefix Language Modeling: }Employed in encoder-decoder and non-causal decoder-only models, this technique uses a non-causal (considering both past and future tokens) prefix for predicting subsequent tokens, offering more flexibility and enhancing the model's adaptability across various language tasks.\\
\textbf{Masked Language Modeling: }Popularized by BERT, this method involves masking certain tokens in the input text and training the model to predict them, helping the model understand word context. An extension, span corruption, masks entire text spans for prediction, further improving contextual comprehension\cite{wang2023language}.
\begin{figure}[htbp]

	\includegraphics[width=1\linewidth]{Figures/pretraining.png}

@@

Fine-tuning has been the most common approach for adapting LLMs to specific tasks. After pre-training on large datasets, these models are further trained on smaller, supervised datasets tailored to the desired task. Fine-tuning updates the weights of the pre-trained model, allowing it to excel at specific tasks like sentiment analysis, machine translation, or question answering. The main advantage of fine-tuning is that it often results in strong performance on task-specific benchmarks. However, it comes with challenges, such as the need for a new large dataset for every task, the risk of overfitting to narrow distributions, and potential poor generalization to out-of-distribution examples\cite{touvron2023llama}.

@@

\section{Popular Models and Datasets}

@@

BERT (Bidirectional Encoder Representations from Transformers) is a groundbreaking language representation model that pretrains deep bidirectional representations from unlabeled text by jointly conditioning on both left and right contexts in all layers. This bidirectional approach enables BERT to be fine-tuned with just a single additional output layer for various tasks, such as question answering and language inference, without the need for extensive architectural modifications. BERT has achieved remarkable performance, setting new state-of-the-art results across eleven NLP tasks, including a GLUE score of 80.5\%, MultiNLI accuracy of 86.7\%, SQuAD v1.1 F1 score of 93.2\%, and SQuAD v2.0 F1 score of 83.1\%. Its combination of conceptual simplicity and empirical effectiveness makes BERT a powerful and versatile tool for a wide range of natural language processing applications\cite{devlin2019bert}.
\subsection{T5}
The T5 (Text-To-Text Transfer Transformer) model represents a unified framework for natural language processing (NLP), designed to address various text processing tasks by treating them as "text-to-text" problems. This approach involves converting all tasks into a format where both input and output are text, allowing the same model, objective, and training procedure to be applied across different tasks. T5 leverages extensive pre-training on a large, unlabeled dataset, enabling it to develop general-purpose knowledge that enhances its performance on a range of NLP tasks, such as question answering, document summarization, and sentiment classification\cite{raffel2023exploring}.
\subsection{LLAMA2}
\subsection{WebText }
WebText is a dataset developed by OpenAI to encapsulate a wide variety of high-quality web content. It comprises text sourced from outbound links on popular websites like Reddit, chosen based on their popularity and the quality of their content. Unlike Common Crawl, which indiscriminately includes a broad range of web pages, WebText is more selective, focusing on high-quality, engaging material. Although smaller in size compared to Common Crawl, WebText is curated to ensure the text reflects diverse, well-written content, making it particularly valuable for training language models to produce coherent and contextually appropriate text\cite{radford2019language}.
\subsection{RuleTaker}
RuleTaker is a collection of datasets designed to assess the deductive reasoning capabilities of language models. Each dataset includes a set of facts, rules, and a boolean question that the model must use logical reasoning to answer. The datasets feature synthetic subsets requiring various levels of reasoning complexity, with different numbers of deduction steps necessary to reach an answer. Among the datasets are the Bird dataset, which illustrates McCarthy’s abnormality problem , the Electricity dataset, which models appliance functions, and the ParaRules corpus, where sentences like “Bob is cold” are paraphrased to “In the snow sits Bob, crying from being cold.” The collection comprises 580,000 training examples, 84,000 validation examples, and 173,000 testing examples\cite{helwe2024}.
\begin{figure}
	\includegraphics[width=0.7\linewidth]{Figures/dataset-exmple.png}
	\caption{exammple of RuleTaker dataset }
\section{Challenges and Limitations}
LLMs have greatly advanced NLP, but they also present challenges such as high computational costs, issues with adversarial robustness, and difficulties in interpretability. As these models scale, they encounter new challenges in scalability, privacy, and real-time processing. Ongoing research is exploring multi-modality, transfer learning, and continuous learning, which bring additional complexities and highlight the evolving impact of LLMs in practical applications\cite{Naveed2024}.
\subsection{ Computational Cost}
Training large language models (LLMs) demands significant computational resources, leading to higher production costs and environmental concerns due to the substantial energy used in large-scale training. Although enhancing computational resources can boost performance, the gains diminish over time when the size of the model and dataset stay constant, adhering to the power law of diminishing returns.
\subsection{Overfitting}
Despite their advanced learning abilities, they can overfit to noisy or unusual patterns in their large training datasets, which may result in generating nonsensical responses. The ongoing discussion about Memorization versus Generalization in LLMs revolves around finding an optimal balance. Memorization helps the model retain specific details from its training, allowing it to answer precise questions accurately. On the other hand, generalization enables the model to make predictions and generate responses for new, unseen inputs, which is crucial for handling diverse real-world tasks. The challenge is to find the right balance: excessive memorization can lead to overfitting, reducing the model's flexibility and ability to handle novel inputs.
\subsection{ Interpretability and Explainability}
The "black-box" nature of LLMs makes it difficult to understand how they make decisions, which is vital for gaining broader acceptance and trust, particularly in sensitive areas. Although these models have advanced capabilities, the lack of transparency into their workings limits their effectiveness and reliability. Efforts are underway to enhance the explainability of LLMs to build user trust and promote responsible use of AI. Understanding how LLMs generate their responses is crucial for ensuring they align with human values and legal standards.
\subsection{ Hallucinations}
LLMs sometimes produce "hallucinations," or responses that, despite appearing plausible, are incorrect or do not match the provided information. These hallucinations can be classified into three types:

>>>>

\subsection{Historical development of (NLP)}
The figure below presents a timeline outlining the key developments in Natural Language Processing (NLP)  beginning with rule-based approaches in the 1950s and progressing through the rise of statistical methods and early neural networks in the late 1980s. It then highlights the impact of deep learning from the 2000s onward, leading to the development of pre-trained models like BERT and GPT. The timeline concludes with the emergence of large language models (LLMs) from 2019 to the present, marking a significant shift in NLP research and applications.
\begin{figure}[htbp]
	\centerline{\includegraphics[scale=.7]{Figures/the_historyOf_nlp.png}}
	\caption{The Vauquois triangle, illustrating the foundations of machine translation.}
	\label{the_historyOf_nlp.png}
\end{figure}

@@

\subsection{Basics Concepts in Neural Networks}
\textbf{Weights}

Variables on the edges between nodes that are multiplied with node outputs to form the input for the next layer.Weights are crucial for training and tuning a neural network and are often initialized within a range of -1 to 1.  \\
\textbf{Biases}

Additional nodes in hidden and output layers that connect to every node within their respective layers but not to the previous layer. Biases add a constant value (typically 1 or -1) to the input of a layer, helping shift the activation function and aiding in effective learning. 
\textbf{Activation Functions}

These functions introduce non-linearity to the network, enabling it to learn and model complex data patterns. Common activation functions include the sigmoid, hyperbolic tangent (tanh), and rectified linear unit (ReLU). \cite{taylor2017neural}.

\subsection{ Recurrent Neural Networks (RNNs)}
Recurrent Neural Networks (RNNs) are a type of neural networks specifically designed for modeling and processing sequential data, such as text or time series. They are particularly well-suited for handling inputs of variable length by maintaining a dynamic internal state. An RNN comprises a hidden state $h$ and an optional output $y$, and operates over an input sequence $x = (x_1, x_2, \dots, x_t)$.

At each time step $t$, the hidden state $h^{(t)}$ is updated based on the previous hidden state $h^{(t-1)}$ and the current input $x_t$ using a non-linear transition function $f$, as shown in Equation~\ref{eq:rnn_update}:

\begin{equation}
	h^{(t)} = f(h^{(t-1)}, x_t)
	\label{eq:rnn_update}
\end{equation}

The function $f$ can be a simple non-linear activation such as the logistic sigmoid or hyperbolic tangent, or a more complex function such as a Long Short-Term Memory (LSTM) unit or a Gated Recurrent Unit (GRU).
\begin{figure}[htbp]

	\centerline{\includegraphics[width=1\linewidth]{Figures/RNN.png}}
	\caption{RNN Architecture.}
	\label{RNN.png}
\end{figure}

This architecture enables RNNs to retain information from previous time steps, allowing the model to capture temporal dependencies and patterns that span across various positions in the sequence. Furthermore, the use of shared parameters across time steps facilitates generalization over sequences of different lengths and enhances the network's ability to learn long-range dependencies.

Despite known challenges, such as difficulties in capturing long-term dependencies due to issues like vanishing gradients, RNNs remain highly effective for tasks that require sequential reasoning or contextual awareness. As a result, they are widely used in applications ranging from time series forecasting to natural language processing.
\cite{cho2014learning}

\subsection{ Long-Short Term Memory (LSTM)}
Long Short-Term Memory (LSTM) networks, designed by Hochreiter and Schmidhuber, are an improved version of recurrent neural networks (RNNs) designed to learn long-term dependencies in sequential data, making them suitable for tasks like time series forecasting and language translation
They address the limitations of traditional RNNs by introducing a memory cell that maintains information over extended periods. This memory cell is regulated by three gates—input, forget, and output gates—which control the flow of information in and out of the cell as shown in the fig \cite{geeksforgeeks_lstm}
\begin{figure}[htbp]
	\centering\includegraphics[width=0.9\linewidth]{Figures/LSTM.png}
	\caption{LSTM Architecture.}
	\label{/LSTM.png}
\end{figure}

Forget Gate: Decides what information to discard from the cell state
% Forget Gate
\begin{equation}
	f_t = \sigma \left( W_f \cdot [h_{t-1}, x_t] + b_f \right)
\end{equation}

% Input Gate: Determines what new information to add to the cell state
\begin{equation}
	i_t = \sigma \left( W_i \cdot [h_{t-1}, x_t] + b_i \right)
\end{equation}
\begin{equation}
	\tilde{C}_t = \tanh \left( W_c \cdot [h_{t-1}, x_t] + b_c \right)
\end{equation}

% Cell state update (commonly used but missing in your snippet)
\begin{equation}
	C_t = f_t \odot C_{t-1} + i_t \odot \tilde{C}_t
\end{equation}

% Output Gate: Controls what information to output from the cell state
\begin{equation}
	o_t = \sigma \left( W_o \cdot [h_{t-1}, x_t] + b_o \right)
\end{equation}
\begin{equation}
	h_t = o_t \odot \tanh(C_t)
\end{equation}

It is worth mentioning that Bidirectional LSTM networks extend the traditional LSTM architecture by processing data in both forward and backward directions. This approach enables the network to capture dependencies from both past and future contexts, improving its ability to resolve temporal dependencies.

Bidirectional LSTMs are particularly effective at handling multidimensional problems, encapsulating spatially and temporally distributed information, and dealing with incomplete data through flexible connection mechanisms \cite{hochreiter1997long}. 
\section{ Transformer Architecture}
The Transformer architecture is a deep learning model introduced in June 2017 by Vaswani et al. from Google Brain. Their paper, titled "Attention Is All You Need," presented a groundbreaking approach to processing sequential data through the use of a self-attention mechanism. This innovative method allows the model to assign different levels of importance to various parts of the input, enabling it to capture long-range dependencies much more effectively than earlier models like RNNs and LSTMs.The original Transformer model is structured as a stack of six layers, where the output of each layer i serves as the input to the subsequent layer i+1, continuing this process until the final prediction is reached.It features a six-layer encoder on the left and a corresponding six-layer decoder on the right,  both of which work together to transform input sequences into meaningful outputs. Each encoder and decoder consists of six identical layers that allow the model to process and generate language efficiently \cite{rothman2021transformers}.
\begin{figure}[htbp]

@@

\end{figure}
\subsection{Encoder and Decoder Stacks}
\textbf{Encoder}\\ Each layer in the encoder consists of two sub-layers: \\
Multi-head self-attention mechanism: This allows the model to focus on different parts of the input sequence. \\
Position-wise feed-forward network: A fully connected network applied to each position separately and identically. \\
Residual Connections: Residual connections are applied around each sub-layer, followed by layer normalization. The output of each sub-layer is computed as Layer Norm(x + Sublayer(x)). \\
Output Dimension: All sub-layers and embedding layers output vectors of dimension d-model = 512.\\
\textbf{Decoder}\\ Similar to the encoder, the decoder also has 6 identical layers, each containing the same two sub-layers:
Multi-head self-attention mechanism ,
Position-wise feed-forward network. \\
Additional Sub-layer: The decoder includes a third sub-layer that performs multi-head attention over the output of the encoder stack. \\
Residual Connections and Normalization: Like the encoder, residual connections are applied around each sub-layer, followed by layer normalization. \\
Masked Self-Attention: The self-attention mechanism in the decoder is modified to prevent positions from attending to subsequent positions, ensuring that predictions for position i depend only on the known outputs at positions before i \cite{vaswani2017attention}.
\begin{figure}[htbp]

@@

\end{figure}
\subsection{Self-Attention Mechanism in Transformers}

Self-attention is a fundamental component of the Transformer architecture, enabling the model to dynamically assess the relevance of each token in a sequence relative to others. This mechanism facilitates the capture of contextual dependencies and semantic relationships among words.

The self-attention operation involves three key components derived from the input embeddings: the \textit{Query} ($Q$), \textit{Key} ($K$), and \textit{Value} ($V$) matrices. The attention scores are computed by taking the dot product between the query vector of a token and the key vectors of all tokens in the sequence:

\begin{equation}
	\text{Attention}(Q, K, V) = \text{softmax}\left(\frac{QK^\top}{\sqrt{d_k}}\right)V
\end{equation}

Here, $d_k$ denotes the dimension of the key vectors, which is used to scale the dot product to prevent excessively large values that could hinder gradient flow. The softmax function transforms the scaled scores into a probability distribution, which serves as the attention weights. These weights are then applied to the value vectors to compute a weighted sum, yielding the output representation for each token.

This mechanism allows each token to selectively focus on other relevant tokens, thereby enhancing the model’s capacity to understand complex linguistic patterns~\cite{vaswani2017attention}.
\subsection{Positional Encoding:}
Since self-attention mechanisms do not inherently capture the order of words, positional encoding is added to the word embeddings to provide information about the relative position of each word in the sequence.
These encodings are added directly to the input embeddings, allowing the model to consider both the position and the content of 
each word during processing\cite{alammar2018illustratedtransformer}.
\subsection{Role of Multi-Head Attention}
Multi-head attention is a key enhancement of the self-attention mechanism that significantly improves the model’s representational capacity. Instead of relying on a single attention operation, the mechanism projects the input into multiple distinct sets of Query, Key, and Value vectors, referred to as attention heads. Each head independently performs self-attention, allowing the model to capture diverse contextual relationships from different subspaces of the input representation.

The outputs of all attention heads are then concatenated and linearly transformed, producing a unified representation that integrates the complementary information gathered by each head. This mechanism enhances the model’s ability to capture complex patterns across different positions in the sequence, thereby improving overall performance in a variety of natural language understanding tasks\cite{vaswani2017attention}.

\subsection{The Rise of Transformers in Natural Language Processing}
Transformers have revolutionized the field of Natural Language Processing (NLP) by introducing a powerful and efficient way to handle textual data. This advancement has led to the creation of highly effective models like BERT and GPT, with its deep bidirectional context, excels at understanding and improving performance across various NLP tasks. GPT, on the other hand, is renowned for generating coherent and contextually relevant text, significantly advancing applications such as text generation and translation\cite{devlin2019bert}.\\
\textbf{Scalability:} Transformers efficiently handle large datasets and long sequences, overcoming the limitations of RNNs. This scalability allows for training with billions of parameters, enhancing model capabilities.\\
\textbf{Rich Contextual Understanding: }The self-attention mechanism in transformers captures relationships between words across the entire sequence, enabling deep contextual understanding and more accurate language processing.\\
\textbf{Model Efficiency:} Transformers enable parallel processing, which speeds up training and makes them more efficient than sequential models like RNNs. This efficiency supports the rapid development and deployment of advanced language models\cite{vaswani2017attention}.
\section{ Emergence of Large Language Models (LLMs)}

Large Language Models (LLMs) are advanced artificial intelligence systems designed to process and generate human-like text. They are typically based on transformer architectures and are characterized by their enormous scale, with billions of parameters. LLMs are pre-trained on vast amounts of text data in a self-supervised manner, enabling them to develop a broad understanding of language. They are capable of performing a wide range of tasks with minimal task-specific fine-tuning, often achieving significant performance improvements through few-shot or zero-shot learning\cite{brown2020language}.
\subsection{Scaling in Large Language Models (LLMs)}
Scaling is crucial in the evolution of large language models. The history of scaling shows that increasing both model size and dataset size leads to significant improvements in performance across various NLP tasks. For instance, early work by Brants et al. (2007) demonstrated the benefits of using language models trained on vast datasets, such as 2 trillion tokens, which led to significant advancements in machine translation quality. This was followed by efforts like those of Heafield et al. (2013), who scaled traditional models to Web-scale data, and Jozefowicz et al. (2016), who scaled LSTMs to 1 billion parameters, achieving state-of-the-art results on large benchmarks.The advent of transformer-based models marked a significant shift.Models like BERT, GPT-2, and GPT-3, with their enormous parameter counts—up to 175 billion for GPT-3—demonstrated that scaling up not only the model but also the dataset size yields substantial gains in performance. Researchers like Kaplan et al. (2020) and Hoffmann et al. (2022) studied how scaling affects model performance, proposing power laws that show a predictable relationship between model size, dataset size, and performance. These studies emphasized the importance of scaling for the continued progress of LLMs\cite{touvron2023llama}.
\subsection{Pre-training}
Pre-training is a crucial stage in developing Large Language Models (LLMs), where the model learns from extensive unlabeled datasets through a process called self-supervision. This stage allows the model to recognize and internalize a wide range of linguistic patterns, laying the groundwork for fine-tuning on specific tasks.\\
Several pre-training objectives have been implemented to maximize the effectiveness of this learning process, each offering distinct benefits to the model's performance .\\
\textbf{Full Language Modeling:} Used since GPT-2, this approach trains decoder-only models to predict the next token in a sequence based on previous tokens. This autoregressive method enables models like GPT-3 to generate coherent and contextually relevant text.\\
\textbf{Prefix Language Modeling: }Employed in encoder-decoder and non-causal decoder-only models, this technique uses a non-causal (considering both past and future tokens) prefix for predicting subsequent tokens, offering more flexibility and enhancing the model's adaptability across various language tasks.\\
\textbf{Masked Language Modeling: }Popularized by BERT, this method involves masking certain tokens in the input text and training the model to predict them, helping the model understand word context. An extension, span corruption, masks entire text spans for prediction, further improving contextual comprehension\cite{wang2023language}.\\
\textbf{Unified Language Modeling} refers to an approach that integrates multiple training objectives, including causal, non-causal, and masked language modeling. In this framework, masked language modeling differs from traditional bidirectional attention mechanisms by employing unidirectional attention—processing context either from left-to-right or right-to-left, rather than considering both directions simultaneously\cite{dong2019unified}.
\begin{figure}[h]
	\includegraphics[width=0.6\linewidth]{Figures/pretraining.png}

@@

Fine-tuning is a key technique for adapting pre-trained large language models (LLMs) to specific downstream tasks. It encompasses several strategies, each serving different objectives.

One common method is \textbf{transfer learning}, where a pre-trained model is further trained on task-specific data to enhance its performance on targeted applications. This allows models to leverage general language understanding while adapting to more specialized domains.

Another approach is \textbf{instruction tuning}, which involves fine-tuning the model on datasets formatted as instruction-response pairs. These datasets typically include multiple tasks described in natural language, enabling the model to better understand prompts and generalize across various instructions. This method has been particularly effective in improving zero-shot and few-shot performance.

To address ethical and safety concerns, \textbf{alignment tuning} is employed. This involves adjusting the model’s behavior based on human feedback to ensure outputs are helpful, honest, and harmless (the "HHH" criteria). A widely adopted framework for this is reinforcement learning with human feedback (RLHF). RLHF combines reward modeling—where human preferences guide the ranking of responses—with reinforcement learning techniques, such as proximal policy optimization (PPO), to iteratively align model behavior with human values.

These fine-tuning approaches, while powerful, come with trade-offs in terms of data requirements, computational cost, and generalization capabilities. Nonetheless, they remain essential for tailoring LLMs to both functional and ethical requirements across diverse applications.\cite{touvron2023llama}.

@@

\section{Popular Models}

@@

BERT (Bidirectional Encoder Representations from Transformers) is a   language representation model that pretrains deep bidirectional representations from unlabeled text by jointly conditioning on both left and right contexts at every layer. This bidirectional strategy allows BERT to be fine-tuned for diverse downstream tasks, such as question answering and natural language inference, by simply appending a single output layer, thereby eliminating the need for extensive architectural modifications.

BERT has demonstrated exceptional performance, establishing new state-of-the-art results across eleven natural language processing benchmarks. Notably, it achieved a GLUE score of 80.5\%, a MultiNLI accuracy of 86.7\%, a SQuAD v1.1 F1 score of 93.2\%, and a SQuAD v2.0 F1 score of 83.1\%. The model's elegant design, coupled with its empirical success, underscores its versatility and efficacy as a powerful tool for a wide array of natural language processing applications.
\cite{devlin2019bert}.
\subsection{T5}
The T5 (Text-To-Text Transfer Transformer) model represents a unified framework for natural language processing (NLP), designed to address various text processing tasks by treating them as "text-to-text" problems. This approach involves converting all tasks into a format where both input and output are text, allowing the same model, objective, and training procedure to be applied across different tasks. T5 leverages extensive pre-training on a large, unlabeled dataset, enabling it to develop general-purpose knowledge that enhances its performance on a range of NLP tasks, such as question answering, document summarization, and sentiment classification\cite{raffel2023exploring}.
\subsection{LLAMA2}
Llama 2 introduces a new family of transformer-based language models available in both foundation and fine-tuned chat variants. These models incorporate architectural enhancements through optimized training procedures, refined pretraining datasets, and advanced fine-tuning techniques including reinforcement learning from human feedback. Empirical evaluations reveal that Llama 2 outperforms its predecessors and other competitive benchmarks across a range of natural language processing tasks, demonstrating notable improvements in accuracy, safety, and generalization. Its open release not only promotes transparency and reproducibility but also fosters collaborative advancements in the field\cite{touvron2023llama2openfoundation}.

\subsection{Jais}
Jais and Jais-chat are advanced Arabic-centric large language models built on a GPT-3 decoder-only architecture. Pretrained on a diverse dataset of Arabic and English texts including programming language source code—these 13-billion-parameter models excel in Arabic knowledge and reasoning, outperforming existing open Arabic and multilingual models. Notably, they also remain competitive in English tasks despite limited English data. We detail their training, fine-tuning, safety alignment, and evaluation, and release both the foundational Jais and the instruction-tuned Jais-chat variants to foster further research on Arabic LLMs\cite{sengupta2023jaisjaischatarabiccentricfoundation}.

\section{Limitations of Large Language Models}
LLMs have greatly advanced NLP, but they also present challenges As these models scale, they encounter new challenges in scalability, privacy, and real-time processing. Several studies have examined the inherent limitations of large language models (LLMs). Notably, \cite{Bender2021} and \cite{Naveed2024} highlight critical challenges that affect the performance and reliability of these models applications.
\subsection{ Computational Cost}
Training large language models (LLMs) demands significant computational resources, leading to higher production costs and environmental concerns due to the substantial energy used in large-scale training. Although enhancing computational resources can boost performance, the gains diminish over time when the size of the model and dataset stay constant, adhering to the power law of diminishing returns.
\subsection{Bias and Ethical Concerns}
The training data for LLMs often encapsulates various social, cultural, and linguistic biases, which the models can inadvertently learn and propagate. This bias can manifest in outputs that reinforce stereotypes or display partiality, raising significant ethical concerns. Addressing these issues is crucial for ensuring fairness and mitigating unintended consequences in real-world applications.
\subsection{ Hallucinations}
LLMs sometimes produce "hallucinations," or responses that, despite appearing plausible, are incorrect or do not match the provided information. These hallucinations can be classified into three types:

\begin{itemize}
	\item Input-conflicting hallucination: When the model generates responses that do not align with the user's input.
	\item Context-conflicting hallucination: When the model produces content that contradicts information it has previously generated.
	\item Fact-conflicting hallucination: When the model creates responses that conflict with established knowledge.
\end{itemize}
\subsection{Overfitting}
Despite their advanced learning abilities, they can overfit to noisy or unusual patterns in their large training datasets, which may result in generating nonsensical responses. The ongoing discussion about Memorization versus Generalization in LLMs revolves around finding an optimal balance. Memorization helps the model retain specific details from its training, allowing it to answer precise questions accurately. On the other hand, generalization enables the model to make predictions and generate responses for new, unseen inputs, which is crucial for handling diverse real-world tasks. The challenge is to find the right balance: excessive memorization can lead to overfitting, reducing the model's flexibility and ability to handle novel inputs.
